A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-01 paper.tex
Update

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the existance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but, crucially,
where each individual is represented by two nodes (for diploids).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the connectivity information alone
does not tell us which \emph{parts} of their genomes were inherited.

Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that
$c\in N$ is the child node, $p\in N$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure; see
Section XXX),
we traverse the graph rootwards from the leaves.
Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$. The efficiency of this process is discussed
in Section XXX).

>>>>

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. 
Originally, the ARG was defined as a variant 
of the coalescent with recombination stochastic 
process~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral},
but has subsequently come to be seen as a data
structure~\citep{minichiello2006mapping}, i.e., a means of 
describing the \emph{outcome} of some such ancestral process
(the distinction, however, is not clear cut and subfields use the term
differently; see Appendix XXX).
ARG inference was for many years synonymous with estimating realisations
of the coalescent with recombination (with, it must be said, limited success),
and the distinction between process and realisation was not especially important.
The recent introduction of two methods, \relate~\citep{speidel2019method}
and \tsinfer~\citep{kelleher2019inferring}, that estimate structures 
sharing many properties of this classical idea of an ARG, but at previously
unthinkable scale, has led to a surge of interest in the area.
Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
avoid describing the output of the methods as an ARG, the general 
consensus has emerged that they belong to a growing class of 
ARG inference methods~\citep[MORE][]{mathieson2020ancestry},
and we therefore need a more general definition of an ``ARG'' than the classical
Griffiths formulation.

@@

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we provide a  concrete definition of an ARG in terms
of specific genomes and their patterns of inheritance, which
we refer to as a ``genome ARG``, or gARG.
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a ``haplotype'' used by
\cite{shipilina2023origin}; we avoid term haplotype because 
it already has many interpretations, and in particular carries the implication of 
specific mutational patterns [CITES] and the persistance of a particular 
haplotype over many generations. A ``genome'' in our usage corresponds to the 
genetic material present in a particular individual or cell.
We will also use the term ``genome`` in its 
more common sense of ``the genome'' of a species, denoting the full set 
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We consider only patterns of genetic inheritance, with mutational 
processes being entirely separate.

A gARG is then a graph in which nodes represent
monoploid genomes and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs.
\footnote{We use the standard graph theoretic sense of
an ``edge'' here, rather than the somewhat specialised interpretation of \cite{shipilina2023origin}.}
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree, but
where each individual is represented by two nodes (for diploids).
This is illustrated in Fig.~\ref{fig-arg-in-pedigree},
which
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

The edges in a gARG specify that genetic inheritance occurred between an
ancestor and a descendant, but the graph connectivity 
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome 
from an ancestor) we must ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $(0, 2]$ from \textsf{e} and it inherits $(2, 10]$ from \textsf{f}.
These intervals are shown attached to the corresponding graph edges.

% JK: dunno if this will make some readers switch off, but I feel it's 
% important to show that this is a formal model and the concrete definition
% will help some readers?
Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $(\ell, r]$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. To generate the local tree at genome
coordinate $x$, we traverse the graph rootwards from the leaves. 
For concreteness, we can describe this tree as an oriented forest $\pi$
\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
At each node
$u$, we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$.
We then set $\pi_c \leftarrow p$,  and move to the next node $p$.
At the end of this process, $\pi$ fully describes the local tree 
at position $x$.
